---
title: "Data Science Applied to Ag - Final Project - Design"
author: "Charles Bency Appolon"
format:
  html:
    code-fold: false
    embed-resources: true
    toc: true
    number-sections: true
    theme: cerulean
---

# Instructions  
This file contains both the **instructions** for the mid-term project and placeholders for your code. You are required to use this file to produce code, output, and answers to the questions below.  

Besides simply creating output, make sure to interpret the output. You will need to create tables and/or plots to arrive at the answers, and then comment on what you found below it.    

To get you setup, you will need to:  

  - Student #1: create a repository on your GitHub account. You can call this repository "2025dsa_finalproject_groupX", **where X is the number of your group**. Make it public, add a README, add a .gitignore for R.  
  - Student #1: follow the steps we did in class to start a new RStudio project with version control.  
  - Student #1: in your computer, create the sub-folders code, data, output, and move your data set into the `data` folder. Also, student 1 moves this current script into the `code` folder. Do a git stage, commit, push.  
  - Student #1: on GitHub, go the repository settings and invite your partner to be a collaborator in the repository. That will give them push permission.  
  - Now, both students should clone this repository on their computers like we did in class. Make sure this step works well and that you can pull and push from GitHub.  
  - Student 2, after cloning, does a git pull to get all these updates on their computer.  
  - Student 1 and 2 work together to update the README file. README files should explain what the repository is about, the goals of that project, who is working in it, and any other important details you may find.  
  
# Introduction  
Describe here the introduction of your problem. Consider this as a shortened version of your paper, where you will briefly discuss in 3-4 paragraphs what is the issue/gap in literature, and how the data you collected will help answer this gap.  
The persistence of Escherichia coli (E. coli) on fresh produce remains a significant concern for food safety, particularly in the context of biological soil amendments of animal origin (BSAAO). Organic fertilizers, while improving soil health and sustainability, may inadvertently introduce or support the survival of enteric pathogens like E. coli in the agroecosystem. Despite the implementation of recommended pre-harvest intervals, evidence suggests that certain soil conditions or management practices may extend pathogen survival, potentially compromising produce safety.

Current literature has primarily focused on the initial contamination routes and general trends in pathogen decline. However, there remains a gap in understanding how specific soil amendments, environmental conditions, and spatial variability (e.g., location) interact to influence the survival of E. coli. In particular, limited data are available on how amendments like poultry litter or composted materials affect pathogen reduction across time and space, under real-world field conditions.

To address this gap, we conducted a field experiment evaluating E. coli populations in soil amended with different treatments and monitored over time across multiple locations. The study design incorporates key variables such as amendment type, sampling day, and field location to provide a comprehensive view of how these factors contribute to E. coli persistence.

The insights generated from this study can inform both scientific understanding and risk assessment models used by regulatory agencies. Ultimately, they support the development of more refined, evidence-based guidelines for pre-harvest intervals and soil management practices in produce production systems.

# Hypothesis and objectives  
Describe here your hypothesis, followed by your objectives. Make sure your hypothesis are testable and bold, and objectives are clear.  

**Hypothesis**:
The type of soil amendment significantly influences the survival of E. coli in soil, and this effect is modulated by time and location.

**Objectives**:

To evaluate the impact of different soil amendments on E. coli population dynamics over time.

To assess how field location contributes to variability in E. coli survival across amendment types.

To determine whether there is a significant interaction between amendment, location, and sampling day in influencing E. coli persistence in the soil.



# Material and Methods  
Describe here your overall material and methods as it pertains to the analysis you will conduct, including study description, site/setup description, what equipment was used, etc. just like you would in a paper. Make sure to clearly explain what was measured and how.
**Inoculum Preparation**:
A rifampin-resistant E. coli strain (TVS 353) was cultured and prepared to a final concentration of 1.0 × 10⁸ CFU/mL in buffered peptone water (BPW) for field inoculation. Enumeration was confirmed using TSA-R plates.

**Plot Preparation**:
A field study was conducted at two sites in Georgia and Florida from Nov 2022 to Apr 2023. Plots received one of four treatments: poultry litter (PL), heat-treated poultry pellets (HTPP), a positive control (UN), or a negative control (C), with each treatment replicated three times (12 plots total). Soil amendments and inoculum were mixed using a sanitized tiller.

**Plot Inoculation**:
Soil samples were collected on 10 timepoints (days 0 to 140) from the top 5 cm layer. Each sample (~100 g) was composited from five subsamples and processed on the same day.

**Bacterial Enumeration**:
Soil E. coli levels were quantified using spread plating on TSA-R (incubated at 42°C), and MPN was used when counts fell below detection limits. Confirmation was done by streaking MPN cultures on TSA-R.

**Soil Physicochemical Analysis**:
Soil was analyzed at each timepoint for:

Potassium: Extracted with 1 M KCl, shaken, and stored at -30°C.
Ammonia & Nitrate: Measured using an Ammonia Analyzer after KCl extraction.
Electrical Conductivity: Measured with a calibrated conductivity meter in a 1:5 soil-to-water slurry.
pH and Nutrient Analysis: Oven-dried and ground soil was analyzed for nutrients via Mehlich extraction and ICP-OES. pH measurements and C/N analysis were also performed.

## Study design  
Clearly describe your study design here, including treatment design (which factors and levels, the hierarchy among them, etc.), and your experimental design (number of reps/blocks, how was randomization performed, etc.), as we talked about in class.  

The factors were:

Fixed Factors:

Treatment (ammendment): three levels (use of heat-treated poultry pellets, use of poultry litter, and no amendments) applied.

Location (location): Florida and Georgia

Time (Day): Represents the time points at which measurements were taken (0, 1, 3, 7, 14, 28, 56, 84, 112, 140).

Random factor: Plot-id (rep_am) - replication = 3.

A complete randomization design was followed, and treatments were attributed accordingly.

## Statistical analysis  
Describe here your statistical analysis, including what type of ANOVA model you ran (based on the design above), what was your response variable, what were your explanatory variables and how were the explanatory variables treated (random or fixed). Provide your alpha level. Explain which function from which package you used to analyze this data. Explain how you checked linear model assumptions and whether or not they were met. Overall, make sure you explain in sufficient detail that, if given your data, a knowledgeable person would be able to reproduce your analysis exactly.  

In this study, a repeated measures linear mixed effects ANOVA was conducted to examine the effects of amendment types, location, and time (day) on the log-transformed population of E. coli (or MPN) as the response variable. 

The ANOVA model was run using the lm() function and evaluated with the anova() function, with an alpha level of 0.05 to determine statistical significance. If significant factors or interactions were found, post-hoc pairwise comparisons were made using least squares means (LS-means) and Bonferroni adjustment via the emmeans() and cld() functions.

Assumptions Checked:
Normality: Residuals were assessed via Q-Q plots.

Homogeneity of Variance: Checked with residual plots .

Independence: Residuals were checked for autocorrelation, given the repeated measures.

Linearity: Visual checks were done for linear relationships between predictors and the response.

Significant effects were interpreted through pairwise comparisons, with the Bonferroni adjustment applied to control for multiple tests. The analysis was conducted using R packages such as lm(), anova(), emmeans(), and cld() to ensure reproducibility and reliability of results.

# Results  
Here is where the coding is going to happen, and it will be completely up to you. Include under this section as many sub-sections (using ##) and as many chunks needed to create the analytical workflow for your analysis, starting at loading packages and data, wrangling, EDA, modeling, assumptions checking, ANOVA table, means, pairwise comparisons, and final publication-quality plot.  

Make sure to run a model that reflects your study design. Even if your study design does not include one of the designs covered in class, you are still expected to run the most appropriate model. If you need help for references, let me know.  

Before each chunk, describe the steps being performed in that chunk. For example, "Here I will load the data".  

If a chunk produces output, like printing a data frame, statistical summary, a plot, ANOVA table, etc., make sure to write text interpreting what you see and how you can/will use that information to move forward to the next steps in the workflow.  
We are loading our packages here
```{r, warning=FALSE, message=FALSE}
# Load packages
library(readxl)
library(ggplot2)
library(lme4)
library(Matrix)
library(nlme)
library(dplyr)
library(emmeans)
library(multcomp)# for letter display
library(car)
library(readxl)
# Plot diagnostics
library(ggplot2)
library(patchwork)
library(broom.mixed)

```
Here, I will load the data and do some wrangling
```{r, warning=FALSE, message=FALSE}
# Read in the data file
ecoli <- read_excel("../data/FL, GA_soil_ec_bothlocations.xlsx") %>%
  mutate(across(c(day, rep_am, ammendment, location), as.factor))

data <- read_excel("../data/FL,GA_commonnutrient.xls") %>%
  mutate(across(c(rep_no, day, rep_am, ammendment, location), as.factor)) %>%
  left_join(
    ecoli %>%
      dplyr::select(location, day, rep_am, ammendment, count),
    by = c("location", "day", "rep_am", "ammendment")
  ) 
  
summary(data)

# Split data by location
data_fl <- data %>% filter(location == "FL") %>% mutate(ammendment = relevel(ammendment, ref = "PC"))
data_ga <- data %>% filter(location == "GA") %>% mutate(ammendment = relevel(ammendment, ref = "PC"))

```

Here, we do exploratory overview of all the response variable. 
```{r, warning=FALSE, message=FALSE}
# Exploratory Data Analysis (EDA)

# Set theme 
theme_pub <- theme_bw(base_size = 14) +
  theme(panel.grid = element_blank(),
        strip.background = element_rect(fill = "#f0f0f0"),
        legend.position = "top")

# explore
nutrients <- c("pH", "SM", "EC", "Ca", "Mg", "Mn", "Zn", "N", "P", "K", "count")

# Plot nutrient distributions by amendment and location per day
for (nutrient in nutrients) {
  p <- ggplot(data, aes_string(x = "ammendment", y = nutrient, fill = "location")) +
    geom_boxplot(outlier.shape = NA, alpha = 0.8) +
    geom_jitter(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.3, shape = 21) +
    facet_wrap(~day, scales = "free_y") +
    labs(title = paste("Distribution of", nutrient, "by Amendment and Location for each Day"),
         x = "Amendment", y = nutrient) +
    scale_fill_brewer(palette = "Set1") +
    theme_pub
  print(p)
}
```
As described earlier, a LME model with an autoregressive order 1 (AR1) to account for irregular intervals between sampling points will be conducted here.
```{r, warning=FALSE, message=FALSE}
## Let's try some iterations
library(nlme)
library(emmeans)
library(multcompView)

results_list <- list()

for (i in 6:ncol(data)) {
  soil_property <- colnames(data)[i]
  
  cat("Analyzing:", soil_property, "\n")
  
  formula <- as.formula(paste(soil_property, "~ day * ammendment * location"))
  
  m2 <- lme(formula, data = data,
            random = ~ 1 | rep_am, 
            correlation = corAR1(form = ~ 1 | rep_am))
  
  anova_results <- anova(m2)
  cat("ANOVA results for", soil_property, "\n")
  print(anova_results) }
  
if (any(anova_results$`p-value` < 0.05, na.rm = TRUE)) { # Check if any factor is significant
    lsmeans_ammendment <- emmeans(m2, ~ location|day|ammendment)
    summary_lsmeans <- summary(lsmeans_ammendment)
    
    lsmeans_amendment_loc_day <- cld(lsmeans_ammendment, 
                                     reversed = TRUE, 
                                     adjust = "Bonferroni", 
                                     Letters = letters)
    
    results_list[[soil_property]] <- list(
      anova = anova_results,
      lsmeans_summary = summary_lsmeans,
      lsmeans_cld = lsmeans_amendment_loc_day
    )
  }
## Only counts had all the factors being significant

results_list$count

```

Here, we will extract our residuals before concluding any inference from the means priorly extracted. Only E. coli count will be utilized for post-hoc means extraction and comparison.
```{r, warning = FALSE, message = FALSE}
# Residual diagnostics for counts
residuals_df <- augment(m2)


p1 <- ggplot(residuals_df, aes(.fitted, .resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs Fitted", x = "Fitted", y = "Residuals") +
  theme_pub

p2 <- ggplot(residuals_df, aes(sample = .resid)) +
  stat_qq() + stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot", x = "Theoretical", y = "Sample") +
  theme_pub

p3 <- ggplot(residuals_df, aes(.resid)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "white") +
  labs(title = "Histogram of Residuals", x = "Residuals") +
  theme_pub

p1 + p2 + p3
```
The normality, the absence of outliers, the independence of observations, and homoscedaticity of variance are all met for the residuals, as we can observe through the plots. Let's check the normality for the random effect (rep_am).

```{r, warning=FALSE, message=FALSE}
# Fit model only for 'count'
soil_property <- "count"
model <- lme(count ~ day * ammendment * location,
             data = data,
             random = ~1 | rep_am,
             correlation = corAR1(form = ~1 | rep_am))

# Store model in a list
model_list <- list()
model_list[[soil_property]] <- model

# Plot Q-Q for random intercepts
for (soil_property in names(model_list)) {
  model <- model_list[[soil_property]]
  
  ranef_vals <- nlme::ranef(model)
  
  # Convert to data frame
  ranef_df <- data.frame(
    rep_am = rownames(ranef_vals),
    intercept = ranef_vals[, 1]
  )
  
  # Q-Q plot
  p <- ggplot(ranef_df, aes(sample = intercept)) +
    stat_qq() +
    stat_qq_line(color = "red") +
    labs(
      title = paste("Q-Q Plot of Random Effects:", soil_property),
      x = "Theoretical Quantiles",
      y = "Random Intercepts"
    ) +
    theme_minimal()
  
  print(p)
}


```
We can then make inferences, as the normality for random effects seems to be met.
```{r}
results_list$count
```
For instance, the effect of time (day), location and amendment types were found to significantly impact E. coli abundance at 99% confidence level. Therefore, the means will be extracted accordingly.
```{r}
lsmeans_ammendment <- emmeans(m2, ~ location|day|ammendment)
    summary_lsmeans <- summary(lsmeans_ammendment)
    
    lsmeans_amendment_loc_day <- cld(lsmeans_ammendment, 
                                     reversed = TRUE, 
                                     adjust = "Bonferroni", 
                                     Letters = letters) %>%
  as.data.frame() %>%
  mutate(letter=trimws(.group))
    lsmeans_amendment_loc_day
```
Let's make our final plot with the letters
```{r}
#| fig-width: 20
#| fig-height: 16
#| fig-align: "center"
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(viridis)

ggplot(mapping = aes(fill = ammendment)) +
  # Boxplots
  geom_boxplot(data = data,
               aes(x = ammendment, y = count),
               alpha = 0.7,
               color = "black",
               width = 0.6,
               position = position_dodge(width = 0.8),
               outlier.shape = NA) +
  
  # Jittered points
  geom_jitter(data = data,
              aes(x = ammendment, y = count),
              shape = 21,
              size = 2.5,
              stroke = 0.3,
              color = "black",
              alpha = 0.5,
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8)) +

  # CLD letters
  geom_label(data = lsmeans_amendment_loc_day,
             aes(x = ammendment, y = emmean + 0.25, label = letter),
             fill = "white",
             size = 4,
             label.size = 0.2,
             fontface = "bold",
             color = "black",
             inherit.aes = FALSE,
             position = position_dodge(width = 0.8),
             show.legend = FALSE) +
  
  labs(
    x = "Amendment Type",
    y = expression(italic("E. coli") * " population (log CFU/g)")
  ) +
  
  scale_fill_viridis_d(option = "D", begin = 0.2, end = 0.85) +

  # Facet by day (rows) and location (columns)
  facet_grid(day ~ location, scales = "free_y") +

  theme_minimal(base_size = 15) +
  theme(
    panel.grid.major.y = element_line(color = "grey90", size = 0.3),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold", size = 13, color = "black"),
    strip.background = element_rect(fill = "grey90", color = NA),
    panel.spacing = unit(1.5, "lines"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    axis.ticks.y = element_line(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

# Save
ggsave("../output/final_proj.png", width = 20, height = 16, dpi = 300)

```
The results show that the interaction between location, amendment, and time significantly impacted E. coli abundance. Notably, E. coli survived at higher levels in plots amended with HTPP, as confirmed by the significant difference found in Florida and Georgia. These results raise concern about the adequacy of the current National Organic Program's 90-120 day rule. Further risk assessments must be done in order to determine the risk of pathogen transfer to an actual produce, and the final risk to consumers.

# Team work in GitHub  
Whether you are working with your future-self or as duos, make sure to stage, commit, and push after finishing each of the sub-sections above. When committing, write commit messages that are short and descriptive (e.g., finished wrangling).  

If you are working in duos, make sure to split the workload. I will check your collaborative work through the commit history, and if one student has contributed significantly more than the other, than that will impact grades.  

**Tip 1**: to avoid merge conflicts, make sure to **pull** first, and then start working locally. That will ensure that any changes made by your partner will be "downloaded" before you make changes to the files locally.  

**Tip 2**: make use of the Issues on this repository to set up to-do lists and assign tasks to different people. You can also use each issue/task to discuss how things should be run and get to an agreement.  

# Submitting your work  
Once you have developed all the code and answers, make sure to Render this quarto file.  

**Notes on rendering**:  

- Make sure to render your work and inspect how the final html look like.  
- If it does not look professional for whatever reason, then fix the issue, re-render it, recheck.  
- Only send me your work once your html file looks professional.  
- Some potential issues you may encounter and how to fix them:   
  - Some times your code may be creating a table output that is waaay to long and cumbersome to scroll through when rendered. If this is the case, make it more professional looking by using the `head()` function to only print the first handful of rows (instead of thousands of rows).  
  
  - **DO NOT** delete the file's heading levels (# and ##). They set up the proper heading 1 and 2 levels, and I use them to guide my grading.  
  
  - If a given chunk is also outputting warnings or messages, inhibit this behavior by changing the chunk options `message` and `warning` to `FALSE`.  
  
  - If, after rendered, 2 lines of text are connected and you wish to "break line" between them, add 2 extra spaces after the first one.  

After rendering, an .html file will be created on your `code` folder.  

Rename this file to `LASTNAME1-LASTNAME2_finalproject.html`.    
For ex., `Bastos-Mendes_finalproject.html`.

Submit the html file on eLC by **April 30th** 11:59 pm.    













  

  

